---
title: "A 10 minute-ish introduction to linear regression"
author: "Eduardo García-Portugués"
date: "24 Feb 2016"
output: html_document
runtime: shiny
---

Companion script for *"A 10 minute-ish introduction to linear regression"* aimed to illustrate the most basic concepts of simple linear regression.

### Motivating example

Suppose we have a production line where we measured the time (in minutes) required to produce a number of solicited items. There are two random variables: `y` = "time required to produce an order" and `x` = "number of items in the order". We want to explain/predict `y` from `x` from a linear model.

```{r}

# Production time
y <- c(175, 189, 344, 88, 114, 338, 271, 173, 284, 277, 337, 58, 146, 277, 123, 227, 63, 337, 146, 68)

# Number of units in the order
x <- c(195, 215, 243, 162, 185, 231, 234, 166, 253, 196, 220, 168, 207, 225, 169, 215, 147, 230, 208, 172)

# Data scatterplot (pch = 16 for filled points)
plot(y ~ x, pch = 16)

```

### The `lm` function

The linear fit in `R` is done by the `lm` function and the formula `y ~ x`, used to denote that we are interested in regressing `y` over `x`. 

```{r}

# Fit a linear model y = beta0 + beta1 * x
reg <- lm(y ~ x)

# The result is a list with several objects
names(reg)

# The fitted coefficients beta0 (intercept) and beta1 (slope)
reg$coefficients

# The regression line with the minimized distances
plot(y ~ x, pch = 16)
abline(reg, col = 2)
segments(x0 = x, y0 = y, x1 = x, y1 = reg$fitted.values)

```

### The least squares estimate

We can check and visualize that `reg$coefficients` indeed contains the least squares estimates and that they minimize the Residual Sum of Squares (RSS).

```{r}

# Create the design matrix
X <- cbind(1, x)

# The analytical solution
betahat <- solve(t(X) %*% X) %*% t(X) %*% y
betahat

# Minimal RSS
sum((y - X %*% betahat)^2)
```

You can try to get a better RSS. Good luck! :)

```{r, echo = FALSE}

# Define input
inputPanel(
  
  sliderInput("beta0", label = "Intercept:",
              min = -500, max = 300, value = reg$coefficients[1] + 600, step = 5),
  sliderInput("beta1", label = "Slope:",
              min = -3, max = 8, value = 0, step = 0.01)
  
)

# Create plot
renderPlot({
  plot(y ~ x, main = paste("RSS:", round(sum((input$beta0 + input$beta1 * x - y)^2), 1)), pch = 16, xlim = c(100, 300), ylim = c(0, 400))
  abline(a = input$beta0, b = input$beta1, col = 2)
  segments(x0 = x, y0 = y, x1 = x, y1 = input$beta0 + input$beta1 * x)
})
  
```

```{r, echo = FALSE}
# 
#  ### Define a line
#   data_line <- data.frame(
#     x_rng = c(0, 1), 
#     y_rng = c(reg$coefficients[1], reg$coefficients[1] + reg$coefficients[2])
#   )     
#   
# p <- ggvis(data.frame("x" = x, "y" = y), x = ~x, y = ~y)
# layer_points(p) %>%
# layer_model_predictions(model = "lm", stroke := "red") %>%
#     layer_paths(x = ~x_rng, y = ~y_rng, stroke := "blue", data = data_line) %>% 
#   scale_numeric("x", domain = c(140, 270), nice = FALSE, clamp = TRUE) %>% 
#   scale_numeric("y", domain = c(0, 400), nice = FALSE, clamp = TRUE)
# 
```


### Summary of the model

The `summary` function applied to a `lm` object gives the fitted coefficients and its significances ("Pr(>|t|)"), the $R^2$ ("Multiple R-squared") and the fitted error variance ("Residual standard error").
 
```{r}

# Summary of the fit
summary(reg)

```

### Prediction

Prediction of a new observation can be done via the function `predict`, which also provides conficence intervals. The `newdata` argument of `predict` needs a `data.frame`.

```{r}

# Point in which we want a prediction for y
newx <- data.frame(x = 200)

# Prediction with 95% confidence interval
predict(reg, newdata = newx, interval = "prediction", level = 0.95)

# The same prediction
reg$coefficients %*% c(1, 200)

```

### Some words of caution with $R^2$

$R^2$ does not measure the correctness of a linear model but the **usefulness assuming the model is correct**. 

```{r}

# Fixed design
xf <- seq(0, 1, l = 50)

# Errors with different variance
set.seed(123456)
eps1 <- rnorm(50, sd = 0.1)
eps2 <- rnorm(50, sd = 1)

# Responses generated following a linear model
y1 <- 1 + 2 * xf + eps1
y2 <- 1 + 2 * xf + eps2

# Fits
reg1 <- lm(y1 ~ xf)
reg2 <- lm(y2 ~ xf)

# R^2 depends on the 
summary(reg1)
summary(reg2)

# Plot
par(mfrow = c(1, 2))
plot(y1 ~ xf, pch = 16, ylim = c(-1, 5))
abline(a = 1, b = 2, col = 2)
abline(a = reg1$coefficients[1], b = reg1$coefficients[2], col = 3)
plot(y2 ~ xf, pch = 16, ylim = c(-1, 5))
abline(a = 1, b = 2, col = 2)
abline(a = reg2$coefficients[1], b = reg2$coefficients[2], col = 3)

```

A large $R^2$ means *nothing* if the **assumptions of the model do not hold**.

```{r}

# Create data that: 
# 1) does NOT follow a linear model
# 2) the error is heteroskedastic
xf <- seq(0.15, 1, l = 100)
y3 <- 0.2 * sin(2 * pi * xf) + rnorm(n = 100, sd = 0.1 * xf^2)

# Great R^2!?
reg3 <- lm(y3 ~ xf)
summary(reg3)

# But predicting is obviously problematic
plot(y3 ~ xf, pch = 16)
lines(0.2 * sin(2 * pi * xf) ~ xf, col = 2)
abline(reg3, col = 3)

```


